---
title: "Geostatistics - Spatial Correlation and Variograms"
output: html_notebook
---

*Milkias Z. SEMEREAB*

*GEOL0097-Geostatistics, University of Liège*

**[GitHub Link](https://github.com/milkiaszerai/geol-0097-geostatistics)**

*October 31, 2022*

------------------------------------------------------------------------


##### **• Structural Analysis**

Before we dive into variogram and spatial modelling, let's visualize and interpret the distribution and spatial continuity of the variables (heavy metals concentration) in the X, Y and Z directions. 

```{r}
# distribution of Zn values through x-direction?
plot(data$zinc ~ data$x, xlab="Easting", ylab="Zn concentration")
abline(lm(data$zinc ~ data$x), col="red")
```
Note:    
- abline() = add Straight line to a an existing Plot    
- lm() = fits Linear Models (e.g. linear regression)


 
```{r}
#  distribution of Zn values through y-direction?
plot(data$zinc ~ data$y, xlab="Northing", ylab="Zn concentration")
abline(lm(data$zinc ~ data$y), col="red")
```

```{r}
# distribution of Zn values through z-direction?
plot(data$zinc ~ data$elev, xlab="Elevation", ylab="Zn concentration")
abline(lm(data$zinc ~ data$elev), col="red")
```

**Questions:**

- Discuss about the distribution/spatial continuity of Zn values through x, y and z directions?


