---
title: "Geostatistics - Univariate and Bivariate Analysis"
output: html_notebook
---

*Milkias Z. SEMEREAB*

*GEOL0097-Geostatistics, University of Liège*

*September 29, 2022*

------------------------------------------------------------------------

#### **Task 1: Loading the dataset**

**• Create your working directory:** The working directory/working folder is the default location where R will look for files that you want to load and where it will put any files you save. You can Create your working directory in 2 ways:

-   One of the great things about using RStudio Projects is that when you create a new project or open an existing project, it will automatically set your working directory to the appropriate location.

To create a new project in Rstudio: *File \> New Project \> New Directory \> New Project \> Type in the name of the directory to store your project, e.g. "meuse_project"*

-   You can also set the working directory manually using the **setwd()** function.

The first method is highly recommended. Once you create the project, use the **getwd()** function to make sure your default working directory is created.

```{r}
 # Show where your working directory was created:
getwd()
```

**• Load the "meuse dataset"** into the workspace using the **read.csv()** function and store the dataframe as variable called "data".       


```{r}
data = read.csv("meuse_data.csv")
```


*Note:* the meuse dataset should be inside the working directory, otherwise R won't be able to find the meuse CSV (comma seperated values) file and will therefore throw an error.    


#### **Task 2: Quick Exploratory Data Analysis (EDA)** 
To have a look at your dataset, you can use the following methods:

- "Environment” tab in the top right window lists the variables and functions present in the current R session. To display the full dataframe, click on the "data" variable on the Environment tab. 

- When you work with large datasets and data frames, your first task as a geostatistician is to develop a clear understanding of its structure and main elements. Therefore, it is often useful to show only a small part of the entire dataset. The function **head()** enables you to show the first observations of a data frame. Similarly, the function **tail()** prints out the last observations in your dataset.

- Another method that is often used to get a rapid overview of your data is the function **str()** which shows you the structure of your dataset including the total number of observations, the total number of variables, the data type of each variable and the first observations etc..

```{r}
head(data)
tail(data)
str(data)
```
The **meuse dataset** is a popular data set that includes amount of four heavy metals measured in the top soil in a flood plain along the river Meuse. The data set wasintroduced by Burrough and McDonnell, 1998.


**Questions:**

- How many observations (samples or data points) and variables (fields or attributes) the meuse datset have? 

- Which variables contain the spatial data ? (note: spatial = geographical location)  

- How many continuous variables are there? How many categorical variables 
are there?


#### **Task 3: Spatial Data Visualization**

**• Visualization of samples location**
```{r}
plot(data[,1],data[,2],pch=16,cex=0.5, xlab = "Easting(m)", ylab = "Northing(m)",main="Sample Locations",asp=1)
```
The plot() function took the following arguments:

- data[ ,1] and data[ ,2] = the latitude/X and longitude/Y of all the sample points. 
- pch = Plot character. Try to change pch to any other number between 1 and 25. 
- cex = point size
- xlab, ylab and main = to put text label for the x-axis, y-axis and head respectively.
- asp = set the aspect ratio of the plot


**• Visualization of data distribution**    

Visualization of heavy metal concentrations in relation to spatial locations.
```{r}
for(j in 1:4){
plot(data[,1],data[,2],pch=16,cex=2*(data[,j+2]-min(data[,j+2]))/(max
(data[,j+2])-min(data[,j+2])),xlab = "Easting(m)", ylab = "Northing(m)
",main=paste(colnames(data)[j+2],"concentrations (ppm)",sep=" "),asp=1)
}
```

*The process governing heavy metal distribution seems that polluted sediment is carried by the river, and mostly deposited close to the river bank*

**Further Questions:**

- What can you tell about the sampling campaign(s)? 
- Do you have clusters? 
- Can you see concentration anomalies?
- Can you observe any spatial regions with outliers (high heavy metal concentrations)?


**• Other way to visualize the Data using the library *plot3D* ** 

Plot3D is an R package/library containing many functions for 2D and 3D plotting. To install any package in Rstudio:

- Go to "packages" tab > click "install" > type "plot3D" and click install 
- Or use the function **install.packages()**     

```{r}
install.packages("plot3D")
```

Once the package is installed, you have to import/load it to the workspace. You can do this either by:

- Simplying checking the checkbox beside the package name "Plot3D" in the Packages tab 
- Using the library() function to load packages which are already installed. 

```{r}
library(plot3D)
```

**scatter2D()** is a function with in **Plot3D** library that is used to polt a Colored scatter plots with a color variable as points. 

```{r}
scatter2D(data[,1], data[,2],colvar=data[,6],pch = 16, xlab = "Easting(m)", ylab = "Northing(m)",clab="zinc",main="zinc concentrations(ppm)", asp=1)
```

##### **Visualization of categorical Variables (e.g., Soil Type)**

```{r}
scatter2D(data[,1],data[,2],colvar=data[,10],col=c("blue", "green", "brown"),pch=16, xlab="Easting(m)", ylab="Northing(m)", clab=colnames(data)[10],main="Soil Type", asp=1)

# Add legend to the plot using the legend() function
legend("bottomright",legend=c("Soil-1","Soil-2", "Soil-3"), fill=c("blue", "green", "brown"), cex=0.8)
```


#### **Task 4: Univariate EDA**

**• Compute the main descriptive univariate statistics for the 4 heavy metals**

```{r}
summary(data[ ,3:6])
```
**Questions:**  

- Observe the minimum, first quartile, median, third quartile, and maximum heavy metal concentrations in the sample set? 
- Compare the mean and median.


**• Display histograms and boxplots of all four heavy metals concentrations**

```{r}
hist(data$cadmium, col="blue", xlab="Values",main="cadmium concentrations")
hist(data$copper,col="blue", xlab="Values",main="copper concentrations")
hist(data$lead,col="blue", xlab="Values",main="lead concentrations")
hist(data$zinc,col="blue", xlab="Values",main="zinc concentrations")
```

```{r}
boxplot(data$cadmium,col="blue", xlab="Values",main="cadmium concentrations")
boxplot(data$copper,col="blue", xlab="Values",main="copper concentrations")
boxplot(data$lead,col="blue", xlab="Values",main="lead concentrations")
boxplot(data$zinc,col="blue", xlab="Values",main="zinc concentrations")
```


**Questions:** 

- Describe the distribution of these variables. Do they have a symmetric or skewed distribution? 
- For each variable: Does it appear to have more than one population? Does it appear to be any extreme values (“outliers”)?


#### **Task 5: Log-Transform the data**




#### **Task 6: Statistics vs category**

**• Compute summary statistics (e.g. mean) of all four heavy metal concentrations by some categorical variables (e.g. soil type)**

```{r}
aggregate(data[ ,3:6], by=list(data$soil), FUN=mean, na.rm=TRUE)
```
- aggregate() function Splits the data into subsets, computes summary statistics for each, and returns the result in a convenient form.
- by=list() is a list of grouping elements. 
- FUN = a function to compute the summary statistics which can be applied to all data subsets.
- na.rm = TRUE excludes missing values when calculating descriptive statistics in R.

**• Display the histogram and boxplot of log10(Zn) by some categorical variables**

To do this task, you can use  **histogram()** function from the **lattice library** and the already familar **boxplot()** function from the R-Base.  

```{r}
# First install and load the lattice library
library(lattice)

# Histogram of zinc concentrations for each soil type
histogram(~ data$zinc|as.factor(data$soil),col="blue",xlab= "Zinc concerntration")

# boxplot of zinc concentration for flood frequencies 
boxplot(data$zinc ~ data$ffreq, col="blue",xlab="Flood Frequency Class", 
ylab= "Zinc concerntration")
```
**Questions:**

- which soil type is more contaminated with zinc? 
- Is flood frequency associated with zinc concentration ?













